---
title: 'Melanoma Plasticity Paper: scRNAseq plots'
author: "Blake Baleami"
date: "12/16/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## About this file

This is an R markdown file created by Blake with all of Corey's code from the Seurat_v3_SKMEL5_combined_hg38.R file in this same directory that was pulled from the origin branch in GitHub. Any changes to Corey's code will be noted in the comments before an R code chunk. 

# Figure 2

Changes made to the code: 
* comments have been added

```{r}
# load packages
library(Seurat)
library(svglite)
library(ggplot2)
library(dplyr)
library(purrr)
library(ggplot2)
library(cowplot)
library(clusterProfiler)
library(org.Hs.eg.db)
library(reshape2)
library(Hmisc)
library(ggpubr)
library(transport)
```
Blake's notes about this section of code:
`clusterProfiler` requires `BiocManager` to install it. To do this, I used the following command to install the package and let it update and install all associated packages and dependencies (which took awhile):
```BiocManager::install("clusterProfiler")```

`org.Hs.eg.db` also requires `BiocManager` to install it. The process was much faster for this than for `clusterProfiler`.

## Figure 2A

Changes made to the code: 
* comments have been modified and/or added
* changed file path for reading in the data in the first chunk of code to match what is on github
* 

```{r}
# Pull this from the data I shared on Box (data_large)
untreated <- Read10X(data.dir = "../data/Untreated/")
idling <- Read10X(data.dir = "../data/Idling/")

# Initialize the Seurat object with the raw (non-normalized data).
untreated <- CreateSeuratObject(counts = untreated, project = "Untreated", min.cells = 3, min.features = 200)
idling <- CreateSeuratObject(counts = idling, project = "Idling", min.cells = 3, min.features = 200)

```

```{r}
# Add lineage information
## Untreated
cellMeta_untreated <- untreated@meta.data
barcode_cell_df_UT <- read.csv("../data/Untreated_LineageBC_cellBC.csv")
barcode_cell_df_UT <- as.data.frame(t(barcode_cell_df_UT))
colnames(barcode_cell_df_UT) <- c("Barcode", "Cell Barcode")
barcode_cell_UT <- barcode_cell_df_UT[-1,]
barcode_cell_UT$`Cell Barcode` <- paste(barcode_cell_UT$`Cell Barcode`, "-1", sep="")
cellMeta_untreated$lineage <- as.character(barcode_cell_UT$Barcode[match(rownames(cellMeta_untreated),
                                                                         barcode_cell_UT$`Cell Barcode`)])
cellMeta_untreated$lineage[nchar(as.character(cellMeta_untreated$lineage)) != 20] <- "XXX"
lineageMeta_UT <- subset(cellMeta_untreated, select = "lineage")
untreated <- AddMetaData(untreated, lineageMeta_UT, col.name = "lineage")

## Idling
cellMeta_idling <- idling@meta.data
barcode_cell_df_I <- as.data.frame(t(read.csv("../data/Treated_LineageBC_cellBC.csv")))
colnames(barcode_cell_df_I) <- c("Barcode", "Cell Barcode")
barcode_cell_I <- barcode_cell_df_I[-1,]
barcode_cell_I$`Cell Barcode` <- paste(barcode_cell_I$`Cell Barcode`, "-1", sep="")
cellMeta_idling$lineage <- as.character(barcode_cell_I$Barcode[match(rownames(cellMeta_idling),
                                                                     barcode_cell_I$`Cell Barcode`)])
cellMeta_idling$lineage[nchar(as.character(cellMeta_idling$lineage)) != 20] <- "XXX"
lineageMeta_I <- subset(cellMeta_idling, select = "lineage")
idling <- AddMetaData(idling, lineageMeta_I, col.name = "lineage")

combined <- merge(untreated, y = idling, add.cell.ids = c("Untreated", "Idling"), project = "Combined")

combined[["percent.mt"]] <- PercentageFeatureSet(combined, pattern = "^MT-")

combined <- SCTransform(combined, assay = 'RNA',
                        new.assay.name = 'SCT',
                        vars.to.regress = c('percent.mt', 'nFeature_RNA', 'nCount_RNA'))

VlnPlot(combined, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)
```
```{r}
# combined <- subset(combined, subset = nFeature_RNA > 200 & nFeature_RNA < 2500 & percent.mt < 5)
combined <- subset(combined, subset = lineage != "NA")

combined <- NormalizeData(combined)

combined <- CellCycleScoring(object = combined, s.features = cc.genes.updated.2019$s.genes,
                             g2m.features = cc.genes.updated.2019$g2m.genes, assay = 'SCT',
                             set.ident = TRUE)

combined <- SCTransform(combined, assay = 'RNA', new.assay.name = 'SCT',
  vars.to.regress = c('percent.mt', 'nFeature_RNA', 'nCount_RNA', 'S.Score', 'G2M.Score'))

combined <- FindVariableFeatures(combined, assay = 'SCT', selection.method = "vst", nfeatures = 2000)

```
```{r}
# Identify the 10 most highly variable genes
# top10 <- head(VariableFeatures(combined), 10)
# 
# # plot variable features with and without labels
# plot1 <- VariableFeaturePlot(combined)
# plot2 <- LabelPoints(plot = plot1, points = top10, repel = TRUE)
# plot1 + plot2

all.genes <- rownames(combined)
combined <- ScaleData(combined, assay = 'SCT', features = all.genes)

combined <- RunPCA(combined, assay = 'SCT', features = VariableFeatures(object = combined))
VizDimLoadings(combined, dims = 1:2, reduction = "pca")
DimPlot(combined, reduction = "pca")

```
```{r}
combined <- FindNeighbors(combined, assay = 'SCT', dims = 1:10)
combined <- FindClusters(combined, assay = 'SCT', resolution = 0.5)

combined <- RunUMAP(combined, assay = 'SCT', dims = 1:10)

DimPlot(combined, reduction = "umap", group.by = "orig.ident") +
  theme_bw() + scale_color_manual(values = c("blue", "red")) +
  theme(axis.text=element_text(size=14), legend.text = element_text(size=14),
        axis.title=element_text(size=14), legend.position = "none",
        panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  xlim(-11.5, 11.5) + ylim(-11.5, 11.5)
ggsave("../figures/UMAP_combined_SKMEL5_hg38_qcCCReg.svg", width = 4, height = 3)

```
Got an error that you "can't add `ggsave("../figures/UMAP_combined_SKMEL5_hg38_qcCCReg.svg", width = 4, ` to a gg plot object" and a "can't add `   height = 3)` to a ggplot object" when ggsave was included in the DimPlot (previous code had a `+` after the ylim and before the ggsave function). When I removed the plus, the error went away. I haven't worked with .svg files before though, so I searched for how to open them. I used Safari to double check that it saved the plot, and it did, which is what ggsave is supposed to do (it defaults to saving the last plot displayed).

Additionally, this plot looked different than Corey's plot in the manuscript, which is not surprising given UMAP is a stochastic process. We should consider potentially setting a seed if we need to redo any UMAP plots again. Since that is not a plot we need to redo, I didn't focus too much on that right now. 

Other plots that can be made were commented in the code file in Github
```{r}
# DimPlot(combined, reduction = "umap", group.by = "Phase") +
#   theme_bw() + #scale_color_manual(values = c("blue", "red")) +
#   theme(axis.text = element_text(size = 14), legend.position = "right",
#         panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
#   ggsave("UMAP_combined_SKMEL5_hg38_qcCCReg_CCPhase_leg.pdf", width = 5, height = 3)
# 
# DimPlot(combined, reduction = "umap", group.by = "State") +
#   theme_bw() + scale_color_manual(values = c("green3", "gold")) +
#   theme(axis.text = element_text(size = 14), legend.position = "none",
#         panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
#   xlim(-11.5, 11.5) + ylim(-11.5, 11.5) +
#   ggsave("UMAP_combined_SKMEL5_hg38_qcCCReg_CCState_leg.svg", width = 4, height = 3)


# DimPlot(combined, reduction = "umap", group.by = "seurat_clusters") +
#   theme_bw() + #scale_color_manual(values = c("blue", "red")) +
#   theme(axis.text = element_text(size = 14), legend.position = "right",
#         panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
#   ggsave("UMAP_combined_SKMEL5_hg38_qcCCReg_clusters_leg.pdf", width = 5, height = 3)
```


```{r}
######

# Determine metrics to plot present in seurat_control@meta.data
metrics <-  c("nCount_RNA", "nFeature_RNA", "S.Score", "G2M.Score", "percent.mt")

# Extract the UMAP coordinates for each cell and include information about the metrics to plot
qc_data <- FetchData(combined, 
                     vars = c(metrics, "ident", "UMAP_1", "UMAP_2"))

# Adding cluster label to center of cluster on UMAP
umap_label <- FetchData(combined, 
                        vars = c("ident", "UMAP_1", "UMAP_2"))  %>%
  as.data.frame() %>% 
  group_by(ident) %>%
  summarise(x=mean(UMAP_1), y=mean(UMAP_2))


# Plot a UMAP plot for each metric
map(metrics, function(qc){
  ggplot(qc_data,
         aes(UMAP_1, UMAP_2)) +
    theme_bw() +
    geom_point(aes_string(color=qc), 
               alpha = 0.7) +
    scale_color_gradient(guide = "none", 
                         low = "grey90", 
                         high = "blue")  +
    geom_text(data=umap_label, 
              aes(label=ident, x, y)) +
    ggtitle(qc)

}) %>%
  plot_grid(plotlist = .)
  ggsave("../figures/UMAP_combined_SKMEL5_hg38_QCmetrics_qcCCReg.svg",
         width = 9, height = 6)

```
* Upon getting an error message saying it was incorrect to say that the 'guide = FALSE' for the scale_color_gradient and that instead 'guide = "none"' should be used, I changed that line of the code so that the error message no longer displays and to input a more desirable value to the code for no guide
* changed file path for saving figure

```{r}
########

umap_vals <- as.data.frame(Embeddings(combined, reduction = "umap"))
umap_vals_df <- dplyr::bind_rows(umap_vals)
colnames(umap_vals_df) <- c("UMAP_1", "UMAP_2")
pops_all <- subset(combined@meta.data, select = "orig.ident")
umap_vals_df$Population <- unlist(pops_all)
all_Kmeans <- kmeans(umap_vals_df[,1:2,], centers = 3)
umap_vals_df$Kmeans <- all_Kmeans$cluster

ggplot(umap_vals_df, aes(x=UMAP_1, y=UMAP_2, color = factor(Kmeans))) +
  geom_point() + theme_bw() + 
  scale_color_manual(values = c("magenta", "gold", "skyblue"), name = "Clusters") +
  theme(axis.text = element_text(size = 14), legend.position = "right",
        panel.grid.major = element_blank(), panel.grid.minor = element_blank())
ggsave("../figures/combined_clustersSKMEL5.svg", width = 5, height = 3)
```

```{r}
# save(combined, file = "../data/SKMEL5_scRNAseq_hg38_combined.RData")
# load("SKMEL5_scRNAseq_hg38_combined.RData")
```


```{r}
### Get proportion of CC cells in each condition
phase_UT <- FetchData(comb_UT, vars = "Phase")
phase_UT$Condition = "Untreated"
phase_I <- FetchData(comb_I, vars = "Phase")
phase_I$Condition = "Idling"
phase_all <- rbind(phase_UT, phase_I)
phase_all$Condition <- factor(phase_all$Condition, levels = c("Untreated", "Idling"))

# ggplot(phase_all, aes(x = Phase, group = Condition, color = Condition)) +
#   theme_bw() + geom_bar(aes(fill=as.factor(Condition), y = (..count..)), 
#                         position = "dodge", color = "black") +
#   scale_fill_manual(values = c("red", "blue")) +
#   labs(x = "Cell Cycle Phase", y = "Number of Cells") +
#   theme(axis.text.y = element_text(size = 14),
#         legend.position = "right", legend.text = element_text(size = 14),
#         plot.title = element_text(size = 16, hjust = 0.5, face = "bold"), 
#         axis.text=element_text(size=14),
#         legend.title = element_blank(), axis.title=element_text(size=14),
#         panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
#   ggsave("SKMEL5_cellCyclePhaseBreakdown_byCondition.pdf", width = 6, height = 4)
```
I could not find a `comb_UT` or `comb_I` object anywhere, and this is the first reference of it in this script. I'll look through other scripts to see if there is something I'm missing or see if there's another data frame or something that contains this information. 

My workaround for this missing object is to just pull it from the combined data frame in a way that is similar to the formatting that Corey was using. Hopefully this will yield the same results. If not, then I'll continue troubleshooting. 

```{r}
phase_all <- FetchData(combined, vars = c("Phase", "orig.ident"))
colnames(phase_all) <- c("Phase", "Condition")

ggplot(phase_all, aes(x = Phase, group = Condition, color = Condition)) +
  theme_bw() + geom_bar(aes(fill=as.factor(Condition), y = (..count..)),
                        position = "dodge", color = "black") +
  scale_fill_manual(values = c("red", "blue")) +
  labs(x = "Cell Cycle Phase", y = "Number of Cells") +
  theme(axis.text.y = element_text(size = 14),
        legend.position = "right", legend.text = element_text(size = 14),
        plot.title = element_text(size = 16, hjust = 0.5, face = "bold"),
        axis.text=element_text(size=14),
        legend.title = element_blank(), axis.title=element_text(size=14),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank())
```

The first comment and two lines of code are not in Corey's file. It would probably be better for this to be up where the 'phase_all' object was created. 
```{r}
# create phase_UT and phase_I by subsetting from phase_all
phase_UT <- subset(phase_all, Condition == "Untreated")
phase_I <- subset(phase_all, Condition == "Idling")

# Percentages
phase_UT_table <- as.data.frame.matrix(table(phase_UT))
phase_UT_table <- transform(phase_UT_table, percent = Untreated / sum(phase_UT_table$Untreated))
phase_UT_table$Phase <- rownames(phase_UT_table)
phase_UT_table$Condition <- "Untreated"
names(phase_UT_table) <- c("Count", "Percent", "Phase", "Condition")

phase_I_table <- as.data.frame.matrix(table(phase_I))
phase_I_table <- transform(phase_I_table, percent = Idling / sum(phase_I_table$Idling))
phase_I_table$Phase <- rownames(phase_I_table)
phase_I_table$Condition <- "Idling"
names(phase_I_table) <- c("Count", "Percent", "Phase", "Condition")

phase_all_table <- rbind(phase_UT_table, phase_I_table)
phase_all_table_melt <- melt(phase_all_table, id.vars = c("Phase", "Condition"),
                             measure.vars = "Percent")
phase_all_table_melt$Condition <- factor(phase_all_table_melt$Condition, 
                                         levels = c("Untreated", "Idling"))

# ggplot(phase_all_table_melt, aes(x = Phase, y = value,
#                                  group = Condition, fill = Condition)) +
#   theme_bw() + geom_bar(stat = "identity", position = "dodge", color = "black") +
#   scale_fill_manual(values = c("red", "blue")) +
#   scale_y_continuous(labels = scales::percent) +
#   labs(x = "Cell Cycle Phase", y = "Percentage of Cells") +
#   theme(axis.text.y = element_text(size = 14),
#         legend.position = "right", legend.text = element_text(size = 14),
#         plot.title = element_text(size = 16, hjust = 0.5, face = "bold"),
#         axis.text=element_text(size=14),
#         legend.title = element_blank(), axis.title=element_text(size=14),
#         panel.grid.major = element_blank(), panel.grid.minor = element_blank()) 
#   ggsave("../figures/SKMEL5_cellCyclePhaseBreakdown_byCondition_percent.pdf", width = 6, height = 4)

# ggplot(phase_all_table_melt, aes(x = Condition, y = value,
#                                  group = Phase, fill = Phase)) +
#   theme_bw() + geom_bar(stat = "identity", position = "fill", color = "black") +
#   # scale_fill_manual(values = c("red", "blue")) +
#   scale_y_continuous(labels = scales::percent) +
#   labs(x = "Condition", y = "Percentage of Cells") +
#   theme(axis.text.y = element_text(size = 14),
#         legend.position = "right", legend.text = element_text(size = 14),
#         plot.title = element_text(size = 16, hjust = 0.5, face = "bold"),
#         axis.text=element_text(size=14),
#         legend.title = element_blank(), axis.title=element_text(size=14),
#         panel.grid.major = element_blank(), panel.grid.minor = element_blank())
#   ggsave("../figures/SKMEL5_cellCyclePhaseBreakdown_byPhase_percent.pdf", width = 6, height = 4)
```

For the following code chunk, I could not do a similar workaround as before since the 'combined' seurat object did not have a "State" variable. Therefore, I used the previous 'phase_all' data frame to convert the phase to a state, where any cell in G1 is "nondividing" and any cell in S or G2M is "dividing". It isn't the best looking or most efficient code, but it was operational. I commented the code that Corey had since I will likely delete it from my version of the R file in preparation for submitting it to a journal. 
```{r}
### Get proportion of CC cells in each state
# state_UT <- FetchData(comb_UT, vars = "State")
# state_UT$Condition = "Untreated"
# state_I <- FetchData(comb_I, vars = "State")
# state_I$Condition = "Idling"
# state_all <- rbind(state_UT, state_I)
# state_all$Condition <- factor(state_all$Condition, levels = c("Untreated", "Idling"))

cellstate <- rep(NA, 6390) # creating a vector to add as a column to cellstate_all
cellstate_all <- cbind(phase_all, cellstate) # combine the phase_all information with the empty vector

# classify G1 phase cells as "nondividing" and all others as "dividing"
for (cell in 1:nrow(cellstate_all)){
  if (cellstate_all[cell, "Phase"] == "G1"){
    cellstate_all[cell, "cellstate"] = "nondividing"
  }
  else {
    cellstate_all[cell, "cellstate"] = "dividing"
  }
}

# create data frames with just the cell state information
state_all <- cellstate_all[, c(3, 2)]
colnames(state_all) <- c("State", "Condition")
state_UT <- subset(state_all, Condition == "Untreated")
state_I <- subset(state_all, Condition == "Idling")

# ggplot(state_all, aes(x = State, group = Condition, color = Condition)) +
#   theme_bw() + geom_bar(aes(fill=as.factor(Condition), y = (..count..)),
#                         position = "dodge", color = "black") +
#   scale_fill_manual(values = c("red", "blue")) +
#   labs(x = "Cell Cycle State", y = "Number of Cells") +
#   theme(axis.text.y = element_text(size = 14),
#         legend.position = "right", legend.text = element_text(size = 14),
#         plot.title = element_text(size = 16, hjust = 0.5, face = "bold"),
#         axis.text=element_text(size=14),
#         legend.title = element_blank(), axis.title=element_text(size=14),
#         panel.grid.major = element_blank(), panel.grid.minor = element_blank()) 
#   ggsave("../figures/SKMEL5_cellCycleStateBreakdown_byCondition.pdf", width = 6, height = 4)

# Percentages
state_UT_table <- as.data.frame.matrix(table(state_UT))
state_UT_table <- transform(state_UT_table, percent = Untreated / sum(state_UT_table$Untreated))
state_UT_table$State <- rownames(state_UT_table)
state_UT_table$Condition <- "Untreated"
names(state_UT_table) <- c("Count", "Percent", "State", "Condition")

state_I_table <- as.data.frame.matrix(table(state_I))
state_I_table <- transform(state_I_table, percent = Idling / sum(state_I_table$Idling))
state_I_table$State <- rownames(state_I_table)
state_I_table$Condition <- "Idling"
names(state_I_table) <- c("Count", "Percent", "State", "Condition")

state_all_table <- rbind(state_UT_table, state_I_table)
state_all_table_melt <- melt(state_all_table, id.vars = c("State", "Condition"),
                             measure.vars = "Percent")
state_all_table_melt$Condition <- factor(state_all_table_melt$Condition, 
                                         levels = c("Untreated", "Idling"))

# ggplot(state_all_table_melt, aes(x = State, y = value,
#                                  group = Condition, fill = Condition)) +
#   theme_bw() + geom_bar(stat = "identity", position = "dodge", color = "black") +
#   scale_fill_manual(values = c("red", "blue")) +
#   scale_y_continuous(labels = scales::percent) +
#   labs(x = "Cell Cycle State", y = "Percentage of Cells") +
#   theme(axis.text.y = element_text(size = 14),
#         legend.position = "right", legend.text = element_text(size = 14),
#         plot.title = element_text(size = 16, hjust = 0.5, face = "bold"),
#         axis.text=element_text(size=14),
#         legend.title = element_blank(), axis.title=element_text(size=14),
#         panel.grid.major = element_blank(), panel.grid.minor = element_blank()) 
#   ggsave("SKMEL5_cellCycleStateBreakdown_byCondition_percent.pdf", width = 6, height = 4)

# ggplot(state_all_table_melt, aes(x = Condition, y = value,
#                                  group = State, fill = State)) +
#   theme_bw() + geom_bar(stat = "identity", position = "fill", color = "black") +
#   scale_fill_manual(values = c("green3", "gold")) +
#   scale_y_continuous(labels = scales::percent) +
#   labs(x = "Condition", y = "Percentage of Cells") +
#   theme(axis.text.y = element_text(size = 14),
#         legend.position = "none", legend.text = element_text(size = 14),
#         plot.title = element_text(size = 16, hjust = 0.5, face = "bold"),
#         axis.text=element_text(size=14),
#         legend.title = element_blank(), axis.title=element_text(size=14),
#         panel.grid.major = element_blank(), panel.grid.minor = element_blank()) 
#   ggsave("../figures/SKMEL5_cellCycleStateBreakdown_byState_percent.pdf", width = 2.5, height = 4)

```

Corey's code did not have the argument `base = exp(1)` in the `FindMarkers` function. A recent update has made the new default a log base 2 instead of natural log. For the consistency of our analysis, I included an argument that defaults back to the natural log as was done in the version that Corey used for his analysis. 
```{r}
### do findmarkers (3 comparisons)
Idents(combined) <- "seurat_clusters"

# Idling - cycling vs non-cycling
I_cycling.markers <- FindMarkers(object = combined, ident.1 = 6,
                                  ident.2 = c(0,2,4), min.pct = 0.25, base = exp(1))
I_noncycling.markers <- FindMarkers(object = combined, ident.1 = c(0,2,4),
                                 ident.2 = 6, min.pct = 0.25, base = exp(1))

I_cycling_GO <- enrichGO(gene = rownames(subset(I_cycling.markers, avg_logFC > 0.5)),
                         universe = row.names(combined@assays$RNA@data),
                         OrgDb = org.Hs.eg.db,
                         keyType = 'SYMBOL',
                         ont = "BP",
                         pAdjustMethod = "BH",
                         pvalueCutoff  = 0.01,
                         qvalueCutoff  = 0.05)

```

I could not get this plot to work for the GO analysis. At first, `Lattice` was masking this from `clusterProfiler`. I then found the function in enrichplot, which was already loaded. So I pushed the function through that package. It needed an "orderBy" parameter, so I looked up some documentation to find that enrichResult plots use "GeneRatio" in that argument. Then I kept getting a "Error in ans[ypos] <- rep(yes, length.out = len)[ypos] : 
  replacement has length zero", which I could not find many solutions for online other than changing the  pvalueCutoff, which would make the analysis different from Corey's. Additionally, it looks like some GO terms were pulled out of that analysis when I look at the I_cycling_GO object and what it is storing, so there should be something to plot. I'm not sure what it is trying to draw from that is an empty vector. 
```{r}
enrichplot::dotplot(I_cycling_GO,  showCategory=10, orderBy = "GeneRatio") 
ggsave("../figures/Idling_smallClusterGO_BP.pdf", width = 6, height = 4)

I_noncycling_GO <- enrichGO(gene = rownames(subset(I_noncycling.markers, avg_logFC > 0.5)),
                         universe = row.names(combined@assays$RNA@data),
                         OrgDb = org.Hs.eg.db,
                         keyType = 'SYMBOL',
                         ont = "CC",
                         pAdjustMethod = "BH",
                         pvalueCutoff  = 0.01,
                         qvalueCutoff  = 0.05)

enrichplot::dotplot(I_noncycling_GO, showCategory=20, orderBy = "GeneRatio")
```

None of the I_markers avg_logFC or avg_log2FC (when omitting the 'base = exp(1)' argument) > 0.5, so I cannot get the enrichGO analysis to run. 
```{r}
# Idling vs Untreated
I_markers <- FindMarkers(object = combined, ident.1 = c(0,2,4,6),
                                 ident.2 = c(1,3,5,7,8), min.pct = 0.25, base = exp(1))
UT_markers <- FindMarkers(object = combined, ident.1 = c(1,3,5,7,8),
                                  ident.2 = c(0,2,4,6), min.pct = 0.25)

I_GO <- enrichGO(gene = rownames(subset(I_markers, avg_logFC > 0.5)),
                            universe = row.names(combined@assays$RNA@data),
                            OrgDb = org.Hs.eg.db,
                            keyType = 'SYMBOL',
                            ont = "BP",
                            pAdjustMethod = "BH",
                            pvalueCutoff  = 0.01,
                            qvalueCutoff  = 0.05)

enrichplot::dotplot(I_GO, showCategory=10) 
ggsave("../figures/IdlingGO_BP.pdf", width = 8, height = 4)

UT_GO <- enrichGO(gene = rownames(subset(UT_markers, avg_log2FC > 0.5)),
                 universe = row.names(combined@assays$RNA@data),
                 OrgDb = org.Hs.eg.db,
                 keyType = 'SYMBOL',
                 ont = "CC",
                 pAdjustMethod = "BH",
                 pvalueCutoff  = 0.01,
                 qvalueCutoff  = 0.05)

enrichplot::dotplot(UT_GO, showCategory=20, orderBy = "GeneRatio")

# Untreated - outcast vs rest
UT_outcast.markers <- FindMarkers(object = combined, ident.1 = 7,
                         ident.2 = c(1,3,5,8), min.pct = 0.25)
UT_rest.markers <- FindMarkers(object = combined, ident.1 = c(1,3,5,8),
                          ident.2 = 7, min.pct = 0.25)

UT_outcast_GO <- enrichGO(gene = rownames(subset(UT_outcast.markers, avg_logFC > 0.5)),
                  universe = row.names(combined@assays$RNA@data),
                  OrgDb = org.Hs.eg.db,
                  keyType = 'SYMBOL',
                  ont = "BP",
                  pAdjustMethod = "BH",
                  pvalueCutoff  = 0.01,
                  qvalueCutoff  = 0.05)

dotplot(UT_outcast_GO, showCategory=10, orderBy = "GeneRatio") 
ggsave("../figures/Untreated_smallClusterGO_BP.pdf", width = 9, height = 4)

UT_rest_GO <- enrichGO(gene = rownames(subset(UT_rest.markers, avg_logFC > 0.5)),
                          universe = row.names(combined@assays$RNA@data),
                          OrgDb = org.Hs.eg.db,
                          keyType = 'SYMBOL',
                          ont = "CC",
                          pAdjustMethod = "BH",
                          pvalueCutoff  = 0.01,
                          qvalueCutoff  = 0.05)

enrichplot::dotplot(UT_rest_GO, showCategory=20, orderBy = "GeneRatio")

# Untreated vs Idling - Small
I_small.markers <- FindMarkers(object = combined, ident.1 = 6,
                                  ident.2 = 7, min.pct = 0.25)
UT_small.markers <- FindMarkers(object = combined, ident.1 = 7,
                               ident.2 = 6, min.pct = 0.25)

I_small_GO <- enrichGO(gene = rownames(subset(I_small.markers, avg_logFC > 0.5)),
                          universe = row.names(combined@assays$RNA@data),
                          OrgDb = org.Hs.eg.db,
                          keyType = 'SYMBOL',
                          ont = "CC",
                          pAdjustMethod = "BH",
                          pvalueCutoff  = 0.01,
                          qvalueCutoff  = 0.05)

enrichplot::dotplot(I_small_GO, showCategory=20, orderBy = "GeneRatio")

UT_small_GO <- enrichGO(gene = rownames(subset(UT_small.markers, avg_logFC > 0.5)),
                       universe = row.names(combined@assays$RNA@data),
                       OrgDb = org.Hs.eg.db,
                       keyType = 'SYMBOL',
                       ont = "CC",
                       pAdjustMethod = "BH",
                       pvalueCutoff  = 0.01,
                       qvalueCutoff  = 0.05)

enrichplot::dotplot(UT_small_GO, showCategory=20, orderBy = "GeneRatio")

# Untreated vs Idling - Large
I_large.markers <- FindMarkers(object = combined, ident.1 = c(0,2,4),
                               ident.2 = c(1,3,5,8), min.pct = 0.25)
UT_large.markers <- FindMarkers(object = combined, ident.1 = c(1,3,5,8),
                                ident.2 = c(0,2,4), min.pct = 0.25)

I_large_GO <- enrichGO(gene = rownames(subset(I_large.markers, avg_logFC > 0.5)),
                       universe = row.names(combined@assays$RNA@data),
                       OrgDb = org.Hs.eg.db,
                       keyType = 'SYMBOL',
                       ont = "CC",
                       pAdjustMethod = "BH",
                       pvalueCutoff  = 0.01,
                       qvalueCutoff  = 0.05)

enrichplot::dotplot(I_large_GO, showCategory=20, orderBy = "GeneRatio")

UT_large_GO <- enrichGO(gene = rownames(subset(UT_large.markers, avg_logFC > 0.5)),
                        universe = row.names(combined@assays$RNA@data),
                        OrgDb = org.Hs.eg.db,
                        keyType = 'SYMBOL',
                        ont = "CC",
                        pAdjustMethod = "BH",
                        pvalueCutoff  = 0.01,
                        qvalueCutoff  = 0.05)

enrichplot::dotplot(UT_large_GO, showCategory=20, orderBy = "GeneRatio")


```

Moving on from the GO analysis...

```{r}
### use lineages
# keep_UT <- as.data.frame(sort(table(untreated@meta.data$lineage), decreasing = TRUE))
# keep_UT <- subset(keep_UT, Var1 != "XXX")
# keep_UT_30 <- subset(keep_UT, Freq >= 30)
# keep_I <- as.data.frame(sort(table(idling@meta.data$lineage), decreasing = TRUE))
# keep_I <- subset(keep_I, Var1 != "XXX")
# keep_I_30 <- subset(keep_I, Freq >= 30)
# bc_30 <- Reduce(intersect, list(keep_UT_30$Var1, keep_I_30$Var1))

# Use top25 untreated lineages
load("../data/top25BCs.RData")
bcs_top25 <- as.character(unique(bcNum_prop_compare_sub$Barcode))
bc_not <- setdiff(combined@meta.data$lineage, bcs_top25)
test <- combined@meta.data
test$lineageColored[test$lineage %in% bcs_top25] <- test$lineage[test$lineage %in% bcs_top25]
test$lineageColored[test$lineage %in% bc_not] <- "XXX"
lineageColored <- subset(test, select = c("lineageColored"))
combined <- AddMetaData(combined, lineageColored, col.name = "lineageColored")
# combined <- SetAllIdent(object = combined, id = "lineageColored")

cols <- c("CTGAGTCAGAGTGACACACT" = rainbow(25)[1],
          "CTGAGAGTGAGTCTGTCAGT" = rainbow(25)[2],
          "CTGAGTGTGAGAGTGTGTGT" = rainbow(25)[3],
          "CTGACAGTGTCACACAGTGA" = rainbow(25)[4],
          "CTGAGTCACTCACTGAGTGT" = rainbow(25)[5],
          "CTGAGACTCAGACAGACACT" = rainbow(25)[6],
          "CTGAGAGACTCTGTGACTGA" = rainbow(25)[7],
          "CTGACAGACAGTGACTGTCT" = rainbow(25)[8],
          "CTGACTGTCAGACAGAGTGA" = rainbow(25)[9],
          "CTGAGTCAGTCACACTCTGT" = rainbow(25)[10],
          "CTGACAGTGTGTCAGTCTCT" = rainbow(25)[11],
          "CTGAGTGTGACTGTGTGTGA" = rainbow(25)[12],
          "CTGACAGACACTCTCAGTCT" = rainbow(25)[13],
          "CTGACTGTCTGTCAGTGTGT" = rainbow(25)[14],
          "CTGACTGTGTGTCAGTGTGA" = rainbow(25)[15],
          "CTGAGAGTCACTGAGTGTGT" = rainbow(25)[16],
          "CTGAGTGTCACTCTCTCAGA" = rainbow(25)[17],
          "CTGACACACTGTGACTGTGT" = rainbow(25)[18],
          "CTGAGAGTGACAGACTCAGT" = rainbow(25)[19],
          "CTGAGTGTGAGTCTGTGACA" = rainbow(25)[20],
          "CTGACAGTCACACTGACTCA" = rainbow(25)[21],
          "CTGACACACTCTCACTGACA" = rainbow(25)[22],
          "CTGACTCAGTCTGTCTGTCA" = rainbow(25)[23],
          "CTGACTGAGAGTGAGTCACA" = rainbow(25)[24],
          "CTGAGTGACTGTGAGACTGA" = rainbow(25)[25],
          "XXX" = "grey80")

cols1 <- c("CTGAGTCAGAGTGACACACT" = rainbow(25)[1],
          "CTGAGAGTGAGTCTGTCAGT" = "grey80",
          "CTGAGTGTGAGAGTGTGTGT" = "grey80",
          "CTGACAGTGTCACACAGTGA" = "grey80",
          "CTGAGTCACTCACTGAGTGT" = "grey80",
          "CTGAGACTCAGACAGACACT" = "grey80",
          "CTGAGAGACTCTGTGACTGA" = "grey80",
          "CTGACAGACAGTGACTGTCT" = "grey80",
          "CTGACTGTCAGACAGAGTGA" = "grey80",
          "CTGAGTCAGTCACACTCTGT" = "grey80",
          "CTGACAGTGTGTCAGTCTCT" = "grey80",
          "CTGAGTGTGACTGTGTGTGA" = "grey80",
          "CTGACAGACACTCTCAGTCT" = "grey80",
          "CTGACTGTCTGTCAGTGTGT" = "grey80",
          "CTGACTGTGTGTCAGTGTGA" = "grey80",
          "CTGAGAGTCACTGAGTGTGT" = "grey80",
          "CTGAGTGTCACTCTCTCAGA" = "grey80",
          "CTGACACACTGTGACTGTGT" = "grey80",
          "CTGAGAGTGACAGACTCAGT" = "grey80",
          "CTGAGTGTGAGTCTGTGACA" = "grey80",
          "CTGACAGTCACACTGACTCA" = "grey80",
          "CTGACACACTCTCACTGACA" = "grey80",
          "CTGACTCAGTCTGTCTGTCA" = "grey80",
          "CTGACTGAGAGTGAGTCACA" = "grey80",
          "CTGAGTGACTGTGAGACTGA" = "grey80",
          "XXX" = "grey80")

labels = c("1","2","3","4","5","6","7","8","9",
           "10","11","12","13","14","15","16",
           "17","18","19","20","21","22",
           "23","24","25","others")
# DimPlot(combined, reduction = "umap", group.by = "lineageColored") +
#   # geom_text(x=7, y=6, label=sprintf("%d cells", nrow(combined@meta.data))) +
#   scale_color_manual(
#     values = cols,
#     labels = labels
#   ) + theme_bw() +
#   theme(legend.position = "none", legend.title = element_blank(),
#         plot.title = element_text(size = 14, hjust = 0.5, face = "bold"),
#         axis.text=element_text(size=14), legend.text = element_text(size=10),
#         axis.title=element_text(size=14), panel.grid.major = element_blank(),
#         panel.grid.minor = element_blank())
#   ggsave("../figures/umap_combined_lineageID_greater30_legendSpecial.pdf", width = 8, height = 5)

tint <- data.frame(ifelse(combined@meta.data$lineageColored == "XXX", 0.1, 1),
                   row.names = rownames(combined@meta.data))
names(tint) <- "Tint"
combined <- AddMetaData(combined, tint, 'Tint')
# tint1 <- data.frame(ifelse(combined@meta.data$lineageColored == "CTGACAGACACTCTCAGTCT", 1, 0.1),
#                     row.names = rownames(combined@meta.data))
# names(tint1) <- "Tint1"
# combined <- AddMetaData(combined, tint1, 'Tint1')
test_subset <- FetchData(combined, vars = c("UMAP_1", "UMAP_2", "lineage", "S.Score", "G2M.Score",
                                                 "Phase", "old.ident", "lineageColored", "Tint", "Tint1"))

test_subset$lineageColored <- factor(test_subset$lineageColored, levels = unique(bcNum_prop_compare_sub$Barcode))
plt <- ggplot(data = test_subset, aes(x = UMAP_1, y = UMAP_2, color = lineageColored, alpha = Tint)) +
  geom_point() + scale_alpha_continuous(range = c(0.1,1)) + 
  theme_bw() +
  scale_color_manual(values = cols, labels = labels, name = "Barcode") +
  guides(alpha = F) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        axis.text.x = element_text(size = 14, colour = "black"),
        axis.text.y = element_text(size = 14, colour = "black"),
        legend.position = "right", legend.text = element_text(size = 14),
        plot.title = element_text(size = 16, hjust = 0.5, face = "bold"), axis.text=element_text(size=14),
        legend.title = element_text(size=14), axis.title=element_text(size=14)) +
  guides(col = guide_legend(nrow = 13)) #+ ggtitle("CTGACAGACACTCTCAGTCT") +

plt_leg <- ggpubr::get_legend(plt)
as_ggplot(plt_leg) 
ggsave("../figures/umap_combined_lineageID_tinted_legendOnly.svg", width = 2.5, height = 4)

```

```{r}
# ggplot(data = test_subset, aes(x = UMAP_1, y = UMAP_2, color = lineageColored, alpha = Tint)) +
#   geom_point(size = 0.8) + scale_alpha_continuous(range = c(0.1,1)) +
#   theme_bw() +
#   scale_color_manual(values = cols, labels = labels, name = "Barcode") +
#   guides(alpha = "none") +
#   theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
#         axis.text.x = element_text(size = 14, colour = "black"),
#         axis.text.y = element_text(size = 14, colour = "black"),
#         legend.position = "none", legend.text = element_text(size = 14),
#         plot.title = element_text(size = 16, hjust = 0.5, face = "bold"), axis.text=element_text(size=14),
#         legend.title = element_text(size=14), axis.title=element_text(size=14))
#   ggsave("../figures/umap_combined_lineageID_tinted_legendSpecial.svg", width = 3.5, height = 3)


# ggplot(data = test_subset, aes(x = UMAP_1, y = UMAP_2, color = lineageColored, alpha = Tint1)) +
#   geom_point(size = 0.8) + scale_alpha_continuous(range = c(0.1,1)) +
#   theme_bw() +
#   scale_color_manual(values = cols1, labels = labels, name = "Barcode") +
#   guides(alpha = "none") +
#   theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
#         axis.text.x = element_text(size = 14, colour = "black"),
#         axis.text.y = element_text(size = 14, colour = "black"),
#         legend.position = "none", legend.text = element_text(size = 14),
#         plot.title = element_text(size = 16, hjust = 0.5, face = "bold"), axis.text=element_text(size=14),
#         legend.title = element_text(size=14), axis.title=element_text(size=14))
#   ggsave("../figures/umap_combined_lineageID_tinted_legendSpecial.svg", width = 4, height = 3)
```


```{r}
# bcs <- c("CTGACAGACACTCTCAGTCT",
#          "CTGACAGTCACACTGACTCA",
#          "CTGACAGTGTCACACAGTGA",
#          "CTGACAGTGTGTCAGTCTCT",
#          "CTGACTGTCAGACAGAGTGA",
#          "CTGACTGTCTGTCAGTGTGT",
#          "CTGACTGTGTGTCAGTGTGA",
#          "CTGAGACTCAGACAGACACT",
#          "CTGAGAGACTCTGTGACTGA",
#          "CTGAGAGTCACTGAGTGTGT",
#          "CTGAGAGTGACAGACTCAGT",
#          "CTGAGAGTGAGTCTGTCAGT",
#          "CTGAGTCACTCACTGAGTGT",
#          "CTGAGTCAGAGTGACACACT",
#          "CTGAGTCAGTCACACTCTGT",
#          "CTGAGTGTCACTCTCTCAGA",
#          "CTGAGTGTGACTGTGTGTGA",
#          "CTGAGTGTGAGAGTGTGTGT",
#          "CTGAGTGTGACTGTGTGTGA",
#          "CTGAGTGTGAGAGTGTGTGT")

t2 <- as.data.frame(combined@meta.data)
t2$BCnum <- as.integer(factor(t2$lineageColored, levels = unique(bcNum_prop_compare_sub$Barcode)))
BCnum <- subset(t2, select = c("BCnum"))
combined <- AddMetaData(combined, BCnum, col.name = "BCnum")

for (i in seq(25)){
  # print(i)
  tint1 <- data.frame(ifelse(combined@meta.data$BCnum == i, 1, 0.1),
                      row.names = rownames(combined@meta.data))
  names(tint1) <- "Tint1"
  combined <- AddMetaData(combined, tint1, 'Tint1')
  test_subset1 <- FetchData(combined, vars = c("UMAP_1", "UMAP_2", "lineage", "S.Score", "G2M.Score",
                                              "Phase", "old.ident", "lineageColored", "Tint", "Tint1"))
  ggplot() +
    theme_bw() +
    geom_density_2d(data = test_subset1, aes(x = UMAP_1, y = UMAP_2, 
                                             color = old.ident), alpha = 0.4) +
    scale_color_manual(values = c("blue", "red")) +
    geom_point(data = subset(test_subset1, BCnum == i), shape = 21,
               aes(x = UMAP_1, y = UMAP_2, fill = lineageColored),
               size = 0.8, stroke = 0.1) + 
    scale_fill_manual(values = cols, labels = labels) +
    # scale_alpha_continuous(range = c(0.1,1)) +
    # scale_color_manual(values = cols, labels = labels, name = "Barcode") +
    # guides(alpha = "none") +
    theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
          axis.text.x = element_text(size = 12, colour = "black"),
          axis.text.y = element_text(size = 12, colour = "black"),
          legend.position = "none", legend.text = element_text(size = 12),
          plot.title = element_text(size = 12, hjust = 0.5, face = "bold"), axis.text=element_text(size=14),
          legend.title = element_text(size=12), axis.title=element_text(size=12)) +
    ggtitle(paste("Barcode",i)) 
    ggsave(paste0("../figures/UMAP_top25BCs_byNum/umap_combined_lineageID_tinted_legendSpecial_top25_Barcode", i, ".svg"), width = 4, height = 3)
}
```


Now we are getting into the distance stuff. The following looks at the distance from the centroid of the UMAP. I will add a version of this for the PCA space. 

```{r}
####

combined_untreated <- subset(combined, subset = orig.ident == "Untreated")
combined_idling <- subset(combined, subset = orig.ident == "Idling")

UT_UMAP <- data.frame(combined_untreated@reductions$umap@cell.embeddings)
UT_UMAP$UMAP1_dist <- (UT_UMAP$UMAP_1 - colMeans(UT_UMAP)[1])**2
UT_UMAP$UMAP2_dist <- (UT_UMAP$UMAP_2 - colMeans(UT_UMAP)[2])**2
UT_UMAP$dist <- sqrt(UT_UMAP$UMAP1_dist + UT_UMAP$UMAP2_dist)

I_UMAP <- data.frame(combined_idling@reductions$umap@cell.embeddings)
I_UMAP$UMAP1_dist <- (I_UMAP$UMAP_1 - colMeans(I_UMAP)[1])**2
I_UMAP$UMAP2_dist <- (I_UMAP$UMAP_2 - colMeans(I_UMAP)[2])**2
I_UMAP$dist <- sqrt(I_UMAP$UMAP1_dist + I_UMAP$UMAP2_dist)

dist_mat <- data.frame(Condition = rep(c("Untreated", "Idling"), 
                                       times = c(nrow(UT_UMAP), 
                                                 nrow(I_UMAP))),
                       Distance = c(UT_UMAP$dist, I_UMAP$dist))


dist_mat_melt <- melt(dist_mat, id.vars = "Condition", measure.vars = "Distance")
dist_mat_melt.summary <- aggregate(dist_mat_melt$value, by = list(dist_mat_melt$Condition),
                                   FUN = mean)
names(dist_mat_melt.summary) <- c("Condition", "Distance")

dist_mat$Condition <- factor(dist_mat$Condition, levels = c("Untreated", "Idling"))
dist_mat_melt.summary$Condition <- factor(dist_mat_melt.summary$Condition, 
                                          levels = c("Untreated", "Idling"))
ggplot(dist_mat, aes(x=Condition, y=Distance, group = Condition,
                     color = Condition)) + theme_bw() +
  geom_jitter(width = 0.2, alpha = 0.1) +
  geom_crossbar(data = dist_mat_melt.summary, aes(ymin=Distance, ymax=Distance, color = Condition),
                size=0.5, width = 0.5, color = "black") +
  # stat_summary(fun.data=mean_sdl, fun.args = list(mult=1),
  #              geom="errorbar", width=0.2) +
  # stat_summary(fun=mean, geom="point", shape=20, size=5) +
  # ylim(0,5) +
  scale_color_manual(values = c("red", "blue")) +
  labs(x = "Condition", y = "Distance from Centroid") +
  theme(axis.text.y = element_text(size = 14),
        legend.position = "none", legend.text = element_text(size = 14),
        plot.title = element_text(size = 16, hjust = 0.5, face = "bold"),
        axis.text=element_text(size=14),
        legend.title = element_text(size=14), axis.title=element_text(size=14),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank())
#   ggsave("../figures/SKMEL5_conditions_distFromCentroid_byCondition.svg", width = 2.5, height = 5)


```


```{r}
# elbow plot
ElbowPlot(object = combined)
# pull PCA information from seurat object
UT_PCA <- data.frame(combined_untreated@reductions$pca@cell.embeddings)

# new data frame for Untreated PCA distances
UT_PCA_dist <- UT_PCA[, c(1:20)]
# calculate square distance from the mean for each point
UT_PCA_dist$PC1_dist <- (UT_PCA$PC_1 - colMeans(UT_PCA)[1])**2
UT_PCA_dist$PC2_dist <- (UT_PCA$PC_2 - colMeans(UT_PCA)[2])**2
UT_PCA_dist$PC3_dist <- (UT_PCA$PC_3 - colMeans(UT_PCA)[3])**2
UT_PCA_dist$PC4_dist <- (UT_PCA$PC_4 - colMeans(UT_PCA)[4])**2
UT_PCA_dist$PC5_dist <- (UT_PCA$PC_5 - colMeans(UT_PCA)[5])**2
UT_PCA_dist$PC6_dist <- (UT_PCA$PC_6 - colMeans(UT_PCA)[6])**2
UT_PCA_dist$PC7_dist <- (UT_PCA$PC_7 - colMeans(UT_PCA)[7])**2
UT_PCA_dist$PC8_dist <- (UT_PCA$PC_8 - colMeans(UT_PCA)[8])**2
UT_PCA_dist$PC9_dist <- (UT_PCA$PC_9 - colMeans(UT_PCA)[9])**2
UT_PCA_dist$PC10_dist <- (UT_PCA$PC_10 - colMeans(UT_PCA)[10])**2

# calculate square root of the sum of squares for 2 to 10 PCs
UT_PCA_dist$dist_2PC <- sqrt(UT_PCA_dist$PC1_dist + UT_PCA_dist$PC2_dist)
UT_PCA_dist$dist_3PC <- sqrt(UT_PCA_dist$PC1_dist + UT_PCA_dist$PC2_dist + UT_PCA_dist$PC3_dist)
UT_PCA_dist$dist_4PC <- sqrt(UT_PCA_dist$PC1_dist + UT_PCA_dist$PC2_dist + UT_PCA_dist$PC3_dist + UT_PCA_dist$PC4_dist)
UT_PCA_dist$dist_5PC <- sqrt(UT_PCA_dist$PC1_dist + UT_PCA_dist$PC2_dist + UT_PCA_dist$PC3_dist + UT_PCA_dist$PC4_dist + UT_PCA_dist$PC5_dist)
UT_PCA_dist$dist_6PC <- sqrt(UT_PCA_dist$PC1_dist + UT_PCA_dist$PC2_dist + UT_PCA_dist$PC3_dist + UT_PCA_dist$PC4_dist + UT_PCA_dist$PC5_dist + UT_PCA_dist$PC6_dist)
UT_PCA_dist$dist_7PC <- sqrt(UT_PCA_dist$PC1_dist + UT_PCA_dist$PC2_dist + UT_PCA_dist$PC3_dist + UT_PCA_dist$PC4_dist + UT_PCA_dist$PC5_dist + UT_PCA_dist$PC6_dist + UT_PCA_dist$PC7_dist)
UT_PCA_dist$dist_8PC <- sqrt(UT_PCA_dist$PC1_dist + UT_PCA_dist$PC2_dist + UT_PCA_dist$PC3_dist + UT_PCA_dist$PC4_dist + UT_PCA_dist$PC5_dist + UT_PCA_dist$PC6_dist + UT_PCA_dist$PC7_dist + UT_PCA_dist$PC8_dist)
UT_PCA_dist$dist_9PC <- sqrt(UT_PCA_dist$PC1_dist + UT_PCA_dist$PC2_dist + UT_PCA_dist$PC3_dist + UT_PCA_dist$PC4_dist + UT_PCA_dist$PC5_dist + UT_PCA_dist$PC6_dist + UT_PCA_dist$PC7_dist + UT_PCA_dist$PC8_dist + UT_PCA_dist$PC9_dist)
UT_PCA_dist$dist_10PC <- sqrt(UT_PCA_dist$PC1_dist + UT_PCA_dist$PC2_dist + UT_PCA_dist$PC3_dist + UT_PCA_dist$PC4_dist + UT_PCA_dist$PC5_dist + UT_PCA_dist$PC6_dist + UT_PCA_dist$PC7_dist + UT_PCA_dist$PC8_dist + UT_PCA_dist$PC9_dist + UT_PCA_dist$PC10_dist)


# pull PCA information from seurat object
I_PCA <- data.frame(combined_idling@reductions$pca@cell.embeddings)

# new data frame for Idling PCA distances
I_PCA_dist <- I_PCA[, c(1:20)]
# calculate square distance from the mean for each point
I_PCA_dist$PC1_dist <- (I_PCA$PC_1 - colMeans(I_PCA)[1])**2
I_PCA_dist$PC2_dist <- (I_PCA$PC_2 - colMeans(I_PCA)[2])**2
I_PCA_dist$PC3_dist <- (I_PCA$PC_3 - colMeans(I_PCA)[3])**2
I_PCA_dist$PC4_dist <- (I_PCA$PC_4 - colMeans(I_PCA)[4])**2
I_PCA_dist$PC5_dist <- (I_PCA$PC_5 - colMeans(I_PCA)[5])**2
I_PCA_dist$PC6_dist <- (I_PCA$PC_6 - colMeans(I_PCA)[6])**2
I_PCA_dist$PC7_dist <- (I_PCA$PC_7 - colMeans(I_PCA)[7])**2
I_PCA_dist$PC8_dist <- (I_PCA$PC_8 - colMeans(I_PCA)[8])**2
I_PCA_dist$PC9_dist <- (I_PCA$PC_9 - colMeans(I_PCA)[9])**2
I_PCA_dist$PC10_dist <- (I_PCA$PC_10 - colMeans(I_PCA)[10])**2

# calculate square root of the sum of squares for 2 to 10 PCs
I_PCA_dist$dist_2PC <- sqrt(I_PCA_dist$PC1_dist + I_PCA_dist$PC2_dist)
I_PCA_dist$dist_3PC <- sqrt(I_PCA_dist$PC1_dist + I_PCA_dist$PC2_dist + I_PCA_dist$PC3_dist)
I_PCA_dist$dist_4PC <- sqrt(I_PCA_dist$PC1_dist + I_PCA_dist$PC2_dist + I_PCA_dist$PC3_dist + I_PCA_dist$PC4_dist)
I_PCA_dist$dist_5PC <- sqrt(I_PCA_dist$PC1_dist + I_PCA_dist$PC2_dist + I_PCA_dist$PC3_dist + I_PCA_dist$PC4_dist + I_PCA_dist$PC5_dist)
I_PCA_dist$dist_6PC <- sqrt(I_PCA_dist$PC1_dist + I_PCA_dist$PC2_dist + I_PCA_dist$PC3_dist + I_PCA_dist$PC4_dist + I_PCA_dist$PC5_dist + I_PCA_dist$PC6_dist)
I_PCA_dist$dist_7PC <- sqrt(I_PCA_dist$PC1_dist + I_PCA_dist$PC2_dist + I_PCA_dist$PC3_dist + I_PCA_dist$PC4_dist + I_PCA_dist$PC5_dist + I_PCA_dist$PC6_dist + I_PCA_dist$PC7_dist)
I_PCA_dist$dist_8PC <- sqrt(I_PCA_dist$PC1_dist + I_PCA_dist$PC2_dist + I_PCA_dist$PC3_dist + I_PCA_dist$PC4_dist + I_PCA_dist$PC5_dist + I_PCA_dist$PC6_dist + I_PCA_dist$PC7_dist + I_PCA_dist$PC8_dist)
I_PCA_dist$dist_9PC <- sqrt(I_PCA_dist$PC1_dist + I_PCA_dist$PC2_dist + I_PCA_dist$PC3_dist + I_PCA_dist$PC4_dist + I_PCA_dist$PC5_dist + I_PCA_dist$PC6_dist + I_PCA_dist$PC7_dist + I_PCA_dist$PC8_dist + I_PCA_dist$PC9_dist)
I_PCA_dist$dist_10PC <- sqrt(I_PCA_dist$PC1_dist + I_PCA_dist$PC2_dist + I_PCA_dist$PC3_dist + I_PCA_dist$PC4_dist + I_PCA_dist$PC5_dist + I_PCA_dist$PC6_dist + I_PCA_dist$PC7_dist + I_PCA_dist$PC8_dist + I_PCA_dist$PC9_dist + I_PCA_dist$PC10_dist)

condition <- rep(c("Untreated", "Idling"), times = c((nrow(UT_PCA_dist)), (nrow(I_PCA_dist))))
d_test <- c(I_PCA_dist$dist_2PC, UT_PCA_dist$dist_2PC)
distance_PC2 <- c(UT_PCA_dist$dist_2PC, I_PCA_dist$dist_2PC)
distance_PC3 <- c(UT_PCA_dist$dist_3PC, I_PCA_dist$dist_3PC)
distance_PC4 <- c(UT_PCA_dist$dist_4PC, I_PCA_dist$dist_4PC)
distance_PC5 <- c(UT_PCA_dist$dist_5PC, I_PCA_dist$dist_5PC)
distance_PC6 <- c(UT_PCA_dist$dist_6PC, I_PCA_dist$dist_6PC)
distance_PC7 <- c(UT_PCA_dist$dist_7PC, I_PCA_dist$dist_7PC)
distance_PC8 <- c(UT_PCA_dist$dist_8PC, I_PCA_dist$dist_8PC)
distance_PC9 <- c(UT_PCA_dist$dist_9PC, I_PCA_dist$dist_9PC)
distance_PC10 <- c(UT_PCA_dist$dist_10PC, I_PCA_dist$dist_10PC)


dist_mat_PCA <- data.frame(Condition = rep(c("Untreated", "Idling"), times = c((nrow(UT_PCA_dist)), (nrow(I_PCA_dist)))),
                       Distance_PC2 = c(UT_PCA_dist$dist_2PC, I_PCA_dist$dist_2PC), 
                       Distance_PC3 = c(UT_PCA_dist$dist_3PC, I_PCA_dist$dist_3PC), 
                       Distance_PC4 = c(UT_PCA_dist$dist_4PC, I_PCA_dist$dist_4PC), 
                       Distance_PC5 = c(UT_PCA_dist$dist_5PC, I_PCA_dist$dist_5PC), 
                       Distance_PC6 = c(UT_PCA_dist$dist_6PC, I_PCA_dist$dist_6PC), 
                       Distance_PC7 = c(UT_PCA_dist$dist_7PC, I_PCA_dist$dist_7PC),
                       Distance_PC8 = c(UT_PCA_dist$dist_8PC, I_PCA_dist$dist_8PC),
                       Distance_PC9 = c(UT_PCA_dist$dist_9PC, I_PCA_dist$dist_9PC),
                       Distance_PC10 = c(UT_PCA_dist$dist_10PC, I_PCA_dist$dist_10PC))

dist_mat_PCA_melt <- melt(dist_mat_PCA, id.vars = "Condition")
dist_mat_PCA_melt.summary <- aggregate(dist_mat_PCA_melt, by = list(dist_mat_PCA_melt$Condition, dist_mat_PCA_melt$variable),
                                   FUN = mean)[c(1,4)]
names(dist_mat_melt.summary) <- c("Condition", "Distance")

dist_mat$Condition <- factor(dist_mat$Condition, levels = c("Untreated", "Idling"))
dist_mat_melt.summary$Condition <- factor(dist_mat_melt.summary$Condition, 
                                          levels = c("Untreated", "Idling"))

ggplot(dist_mat_PCA, aes(x=Condition, y=Distance_PC2, group = Condition,
                     color = Condition)) + theme_bw() +
  geom_jitter(width = 0.2, alpha = 0.1) +
  geom_crossbar(data = dist_mat_melt.summary, aes(ymin=Distance, ymax=Distance, color = Condition),
                size=0.5, width = 0.5, color = "black") +
  # stat_summary(fun.data=mean_sdl, fun.args = list(mult=1),
  #              geom="errorbar", width=0.2) +
  # stat_summary(fun=mean, geom="point", shape=20, size=5) +
  # ylim(0,5) +
  scale_color_manual(values = c("red", "blue")) +
  labs(x = "Condition", y = "Distance from Centroid") +
  theme(axis.text.y = element_text(size = 14),
        legend.position = "none", legend.text = element_text(size = 14),
        plot.title = element_text(size = 16, hjust = 0.5, face = "bold"),
        axis.text=element_text(size=14),
        legend.title = element_text(size=14), axis.title=element_text(size=14),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank())
#   ggsave("../figures/SKMEL5_conditions_distFromCentroid_byCondition.svg", width = 2.5, height = 5)
```

```{r}
### 

df1 <- as.data.frame(combined@meta.data)
df1$State <-  with(df1, ifelse(Phase %in% c("G2M", "S"), yes="Dividing", no="Nondividing"))
State <- subset(df1, select = c("State"))
combined <- AddMetaData(combined, State, col.name = "State")
save(combined, file = "../data/combined_includingState.RData")
```
This saves the current combined seurat object to this point in an RData file that can be reloaded again instead of having to run all of the code above. 

```{r}
# BC-by-BC Number/Proportion of cells in each CC state
stateBC_df <- FetchData(combined, vars = c("orig.ident", "lineageColored", "State"))
stateBC_df <- subset(stateBC_df, lineageColored != "XXX")
t1 <- stateBC_df
t1$BCnum <- as.integer(factor(t1$lineageColored, levels = unique(bcNum_prop_compare_sub$Barcode)))
t1$lineageColored <- factor(t1$lineageColored, levels = unique(bcNum_prop_compare_sub$Barcode))

# this creates a plot similar to figure 1E but showing the number of cells in the dividing vs nondividing states for untreated cells
ggplot(subset(t1, orig.ident == "Untreated"), 
       aes(x=BCnum, group = State, fill = State)) +
  theme_bw() + geom_bar(stat = "count", position = "dodge", color = "black") +
  scale_fill_manual(values = c("green3", "gold")) +
  labs(x = "Barcode", y = "Number of Cells") +
  theme(axis.text.y = element_text(size = 14),
        legend.position = "right", legend.text = element_text(size = 14),
        plot.title = element_text(size = 16, hjust = 0.5, face = "bold"), 
        axis.text=element_text(size=14),
        legend.title = element_blank(), axis.title=element_text(size=14),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank()) 
# there is a break where it looks like barcode 8 is missing from this plot for some reason

# this produces a similar plot as above, except for idling cells
ggplot(subset(t1, orig.ident == "Idling"), 
       aes(x=BCnum, group = State, fill = State)) +
  theme_bw() + geom_bar(stat = "count", position = "dodge", color = "black") +
  scale_fill_manual(values = c("green3", "gold")) +
  labs(x = "Barcode", y = "Number of Cells") +
  theme(axis.text.y = element_text(size = 14),
        legend.position = "right", legend.text = element_text(size = 14),
        plot.title = element_text(size = 16, hjust = 0.5, face = "bold"), 
        axis.text=element_text(size=14),
        legend.title = element_blank(), axis.title=element_text(size=14),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank()) 

```

There are a lot more nondividing cells in the idling population, which isn't surprising given what was seen in the figures in the manuscript.

```{r}
# Percentages
state_UT_table_byBC <- subset(t1, orig.ident == "Untreated")[,c("State", "BCnum")]
state_UT_byBC <- state_UT_table_byBC %>% 
  dplyr::group_by(State, BCnum) %>% 
  dplyr::summarise (n = n()) %>%
  dplyr::group_by(BCnum) %>%
  dplyr::mutate(freq = n / sum(n))
state_UT_byBC$Condition <- "Untreated"

state_I_table_byBC <- subset(t1, orig.ident == "Idling")[,c("State", "BCnum")]
state_I_byBC <- state_I_table_byBC %>% 
  dplyr::group_by(State, BCnum) %>% 
  dplyr::summarise (n = n()) %>%
  dplyr::group_by(BCnum) %>%
  dplyr::mutate(freq = n / sum(n))
state_I_byBC$Condition <- "Idling"

state_byBC <- as.data.frame(rbind(state_UT_byBC, state_I_byBC))

# ggplot(subset(state_byBC, Condition == "Untreated"), aes(x = as.factor(BCnum), y = freq,
#                        group = State, fill = State)) +
#   theme_bw() + geom_bar(position = position_fill(reverse = TRUE),
#                         stat = "identity", color = "black") +
#   scale_fill_manual(values = c("green3", "gold")) +
#   scale_y_continuous(labels = scales::percent) +
#   labs(x = "Barcode", y = "Percentage of Cells") +
#   ggtitle("Untreated Cell Cycle State") +
#   theme(axis.text.y = element_text(size = 14),
#         legend.position = "bottom", legend.text = element_text(size = 14),
#         plot.title = element_text(size = 16, hjust = 0.5),
#         axis.text=element_text(size=14),
#         legend.title = element_blank(), axis.title=element_text(size=14),
#         panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
#         axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
#   ggsave("../figures/SKMEL5_UT_CellCycleState_proportion.pdf", width = 8, height = 5)

a <- subset(combined@meta.data, orig.ident == "Idling")
a %>%
  dplyr::group_by(State) %>% 
  dplyr::summarise (n = n()) %>%
  dplyr::mutate(prop = n / sum(n))
  
b <- subset(combined@meta.data, orig.ident == "Untreated")
b %>%
  dplyr::group_by(State) %>% 
  dplyr::summarise (n = n()) %>%
  dplyr::mutate(prop = n / sum(n))

# ggplot(subset(state_byBC, Condition == "Idling"), aes(x = as.factor(BCnum), y = freq,
#                                                          group = State, fill = State)) +
#   theme_bw() + geom_bar(position = position_fill(reverse = TRUE),
#                         stat = "identity", color = "black") +
#   scale_fill_manual(values = c("green3", "gold")) +
#   scale_y_continuous(labels = scales::percent) +
#   labs(x = "Barcode", y = "Percentage of Cells") +
#   ggtitle("Idling Cell Cycle State") +
#   theme(axis.text.y = element_text(size = 14),
#         legend.position = "right", legend.text = element_text(size = 14),
#         plot.title = element_text(size = 16, hjust = 0.5),
#         axis.text=element_text(size=14),
#         legend.title = element_blank(), axis.title=element_text(size=14),
#         panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
#         axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
#   ggsave("../figures/SKMEL5_I_CellCycleState_proportion.pdf", width = 8, height = 5)
```
##### THIS FIRST PLOT LOOKS LIKE FIGURE 3B IN VERSION 8 OF THE FIGURES
```{r}

ggplot(subset(state_byBC, State == "Dividing" & Condition == "Untreated"), aes(x = as.factor(BCnum), y = freq)) +
  theme_bw() + geom_bar(stat = "identity", position = "dodge", color = "black", fill = "red") +
  geom_hline(yintercept = 0.670, linetype = 2) +
  scale_y_continuous(labels = scales::percent) +
  labs(x = "Barcode", y = "Percentage of Dividing Cells") +
  theme(axis.text.y = element_text(size = 14),
        legend.position = "right", legend.text = element_text(size = 14),
        plot.title = element_text(size = 16, hjust = 0.5, face = "bold"), 
        axis.text=element_text(size=14),
        legend.title = element_blank(), axis.title=element_text(size=14),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) 
  ggsave("../figures/SKMEL5_UT_CellCycleState_onlyDividing_proportionWAverage.pdf", width = 6, height = 4)

ggplot(subset(state_byBC, State == "Dividing" & Condition == "Idling"), aes(x = as.factor(BCnum), y = freq)) +
  theme_bw() + geom_bar(stat = "identity", position = "dodge", color = "black", fill = "blue") +
  geom_hline(yintercept = 0.166, linetype = 2) +
  scale_y_continuous(labels = scales::percent) +
  labs(x = "Barcode", y = "Percentage of Dividing Cells") +
  theme(axis.text.y = element_text(size = 14),
        legend.position = "right", legend.text = element_text(size = 14),
        plot.title = element_text(size = 16, hjust = 0.5, face = "bold"), 
        axis.text=element_text(size=14),
        legend.title = element_blank(), axis.title=element_text(size=14),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) 
  ggsave("../figures/SKMEL5_I_CellCycleState_onlyDividing_proportionWAverage.pdf", width = 6, height = 4)


```

Couldn't find the state_all_table_byBC_melt object anywhere in Corey's code, so I'm not sure what this used to be and didn't have time yet to investigate. 
```{r}
# ggplot(state_all_table_byBC_melt, aes(x = Condition, y = value,
#                                  group = State, fill = State)) +
#   theme_bw() + geom_bar(stat = "identity", position = "fill", color = "black") +
#   scale_fill_manual(values = c("green3", "orange1")) +
#   scale_y_continuous(labels = scales::percent) +
#   labs(x = "Condition", y = "Percentage of Cells") +
#   theme(axis.text.y = element_text(size = 14),
#         legend.position = "right", legend.text = element_text(size = 14),
#         plot.title = element_text(size = 16, hjust = 0.5, face = "bold"),
#         axis.text=element_text(size=14),
#         legend.title = element_blank(), axis.title=element_text(size=14),
#         panel.grid.major = element_blank(), panel.grid.minor = element_blank())
```


Now we're getting into the clustering that may lead to either the supplementary figure information or figures 2C and 2D. 
```{r}
####
# K-means / Clusters

# these came after the follow code chunk, but test_twoState needs to be defined before that so I moved it up here
DimPlot(combined, reduction = "umap", group.by = "seurat_clusters", label = TRUE)
test_twoState <- FetchData(combined, vars = c("UMAP_1", "UMAP_2", "Phase", "State", "old.ident", "seurat_clusters"))

test_KM <- test_twoState
set.seed(10)
kc <- kmeans(test_KM[,c("UMAP_1", "UMAP_2")], centers = 3)
test_KM$Kclusters <- kc$cluster
head(test_KM)
ggplot(test_KM, aes(x=UMAP_1, y=UMAP_2, color = Kclusters)) +
  geom_point()
# Kclusters <- subset(test_KM, select = c("Kclusters"))
# combined <- AddMetaData(combined, Kclusters, col.name = "Kclusters")
# DimPlot(combined, reduction = "pca", group.by = "Kclusters")
```
#### This looks like figure 2A
```{r}
plt_treat <- ggplot(test_twoState, aes(x = UMAP_1, y = UMAP_2)) +
  geom_point(aes(color = factor(old.ident, levels = c("Untreated", "Idling"))), size = 1) + 
  scale_color_manual(name = "", values = c("red", "blue")) +
  theme_bw() + xlim(-11.5, 11.5) + ylim(-11.5, 11.5) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        axis.text.x = element_text(size = 12, colour = "black"),
        axis.text.y = element_text(size = 12, colour = "black"),
        legend.position = "right", legend.text = element_text(size = 12),
        plot.title = element_text(size = 12, hjust = 0.5, face = "bold"), axis.text=element_text(size=14),
        legend.title = element_text(size=12), axis.title=element_text(size=12)) 
plt_treat  
ggsave("../figures/UMAP_combined_SKMEL5_hg38_qcCCReg_treatmentPoint.svg", width = 4, height = 3)
plt_treat_leg <- ggpubr::get_legend(plt_treat)
as_ggplot(plt_treat_leg) 
ggsave("../figures/UMAP_treatment_legend.svg", width = 2.5, height = 4)

```


#### This looks like figure 2C
```{r}
plt_state <- ggplot(test_twoState, aes(x = UMAP_1, y = UMAP_2)) +
  geom_density_2d(aes(color = factor(old.ident, levels = c("Untreated", "Idling"))), alpha = 0.4) +
  scale_color_manual(name = "", values = c("red", "blue")) +
  geom_point(shape = 21,
             aes(fill = State),
             size = 1, stroke = 0.1) + 
  scale_fill_manual(name = "", values = c("green3", "gold")) +
  theme_bw() + xlim(-11.5, 11.5) + ylim(-11.5, 11.5) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        axis.text.x = element_text(size = 12, colour = "black"),
        axis.text.y = element_text(size = 12, colour = "black"),
        legend.position = "right", legend.text = element_text(size = 12),
        plot.title = element_text(size = 12, hjust = 0.5, face = "bold"), axis.text=element_text(size=14),
        legend.title = element_text(size=12), axis.title=element_text(size=12)) 
plt_state 
ggsave("../figures/UMAP_combined_SKMEL5_hg38_qcCCReg_treatmentDensity_CCStatePoint.svg", width = 4, height = 3)
plt_state_leg <- ggpubr::get_legend(plt_state)
as_ggplot(plt_state_leg) 
ggsave("../figures/UMAP_state_legend.svg", width = 2.5, height = 4)
```
#### Figure 2D Original
```{r}
# the exact clusters were different for me, so I changed the numbers
test_twoState_UT <- subset(test_twoState, old.ident == "Untreated")
test_twoState_UT$Region <- with(test_twoState_UT, 
                                ifelse(seurat_clusters %in% as.factor(c("1","2","6","10")), 
                                       yes="large", no="small"))

test_twoState_I <- subset(test_twoState, old.ident == "Idling")
test_twoState_I$Region <- with(test_twoState_I, 
                             ifelse(seurat_clusters %in% as.factor(c("0","3","4", "7")), 
                                    yes="large", no="small"))

region_UT_table <- test_twoState_UT[,c("State", "Region")]
region_UT_table <- region_UT_table %>% 
  dplyr::group_by(State, Region) %>% 
  dplyr::summarise (n = n()) %>%
  dplyr::group_by(Region) %>%
  dplyr::mutate(freq = n / sum(n))
region_UT_table$Condition <- "Untreated"

region_I_table <- test_twoState_I[,c("State", "Region")]
region_I_table <- region_I_table %>%
  dplyr::group_by(State, Region) %>%
  dplyr::summarise (n = n()) %>%
  dplyr::group_by(Region) %>%
  dplyr::mutate(freq = n / sum(n))
region_I_table$Condition <- "Idling"

# region_all_table_n <- region_all_table %>%
#   dplyr::group_by(name) %>%
#   dplyr::summarise (num = sum(n))

region_all_table <- rbind(region_I_table, region_UT_table)
region_all_table$name <- c("I_L", "I_S", "I_L", "I_S",
                           "UT_L", "UT_S", "UT_L", "UT_S")
region_all_table$name <- factor(region_all_table$name,
                                levels = c("UT_L", "UT_S", "I_L", "I_S"))
# region_all_table$tot <- c()


region_all_table_n <- region_all_table %>%
  dplyr::group_by(name) %>%
  dplyr::summarise (num = sum(n))

region_all_n <- data.frame(name = unique(region_all_table_n$name),
                           label = c(paste("n =", as.character(region_all_table_n$num[1])),
                                     paste("n =", as.character(region_all_table_n$num[2])),
                                     paste("n =", as.character(region_all_table_n$num[3])),
                                     paste("n =", as.character(region_all_table_n$num[4]))))

# ggplot(region_all_table, aes(x = name, y = freq,
#                              group = State, fill = State)) +
#   theme_classic() + geom_bar(stat = "identity", color = "black") +
#   scale_fill_manual(values = c("green3", "gold")) +
#   scale_y_continuous(name = "Percentage of Cells", labels = scales::percent) +
#   scale_x_discrete(name = "Condition",
#                    labels = c(expression(Untreated[large]),
#                               expression(Untreated[small]),
#                               expression(Idling[large]),
#                               expression(Idling[small]))) +
#   # labs(y = "Percentage of Cells") +
#   # ggtitle("Cluster Cell Cycle State") +
#   # geom_text(data = region_all_n,
#   #           aes(name, label, label = label)) +
#   theme(axis.text.y = element_text(size = 14),
#         legend.position = "bottom", legend.text = element_text(size = 14),
#         plot.title = element_text(size = 16, hjust = 0.5),
#         axis.text=element_text(size=14),
#         legend.title = element_blank(), axis.title=element_text(size=14),
#         panel.grid.major = element_blank(), panel.grid.minor = element_blank())
#   ggsave("../figures/SKMEL5_allClusters_CellCycleState_proportion_original.pdf", width = 6, height = 5)
```

#### Figure 2D edited

So the consensus seems to be that there is only 1 larger idling cluster that can be distinct from other clusters of cells (untreated clusters). Therefore, figure 2D is being changed to have idling just be one cluster by itself. 

```{r}
test_twoState_UT <- subset(test_twoState, old.ident == "Untreated")
test_twoState_UT$Region <- with(test_twoState_UT, 
                                ifelse(seurat_clusters %in% as.factor(c("1","2","6","10")), 
                                       yes="large", no="small"))

test_twoState_I <- subset(test_twoState, old.ident == "Idling")
# labeling all idling cells as being in a "large" cluster
test_twoState_I$Region <- with(test_twoState_I, 
                             ifelse(seurat_clusters %in% as.factor(c("0","3","4", "5", "7")),
                                    yes="large", no="large"))

region_UT_table <- test_twoState_UT[,c("State", "Region")]
region_UT_table <- region_UT_table %>% 
  dplyr::group_by(State, Region) %>% 
  dplyr::summarise (n = n()) %>%
  dplyr::group_by(Region) %>%
  dplyr::mutate(freq = n / sum(n))
region_UT_table$Condition <- "Untreated"

region_I_table <- test_twoState_I[,c("State", "Region")]
region_I_table <- region_I_table %>%
  dplyr::group_by(State, Region) %>%
  dplyr::summarise (n = n()) %>%
  dplyr::group_by(Region) %>%
  dplyr::mutate(freq = n / sum(n))
region_I_table$Condition <- "Idling"

# region_all_table_n <- region_all_table %>%
#   dplyr::group_by(name) %>%
#   dplyr::summarise (num = sum(n))

region_all_table <- rbind(region_I_table, region_UT_table)
region_all_table$name <- c("I_L", "I_L",
                           "UT_L", "UT_S", "UT_L", "UT_S")
region_all_table$name <- factor(region_all_table$name,
                                levels = c("UT_L", "UT_S", "I_L"))
# region_all_table$tot <- c()


region_all_table_n <- region_all_table %>%
  dplyr::group_by(name) %>%
  dplyr::summarise (num = sum(n))

region_all_n <- data.frame(name = unique(region_all_table_n$name),
                           label = c(paste("n =", as.character(region_all_table_n$num[1])),
                                     paste("n =", as.character(region_all_table_n$num[2])),
                                     paste("n =", as.character(region_all_table_n$num[3]))))

ggplot(region_all_table, aes(x = name, y = freq,
                             group = State, fill = State)) +
  theme_classic() + geom_bar(stat = "identity", color = "black") +
  scale_fill_manual(values = c("green3", "gold")) +
  scale_y_continuous(name = "Percentage of Cells", labels = scales::percent) +
  scale_x_discrete(name = "Condition",
                   labels = c(expression(Untreated[large]),
                              expression(Untreated[small]),
                              expression(Idling))) +
  # labs(y = "Percentage of Cells") +
  # ggtitle("Cluster Cell Cycle State") +
  # geom_text(data = region_all_n,
  #           aes(name, label, label = label)) +
  geom_text(aes(region_all_n$name[1], 1.05, label = region_all_n$label[1])) + 
  geom_text(aes(region_all_n$name[2], 1.05, label = region_all_n$label[2])) +
  geom_text(aes(region_all_n$name[3], 1.05, label = region_all_n$label[3])) +
  theme(axis.text.y = element_text(size = 14),
        legend.position = "bottom", legend.text = element_text(size = 14),
        plot.title = element_text(size = 16, hjust = 0.5),
        axis.text=element_text(size=14),
        legend.title = element_blank(), axis.title=element_text(size=14),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank())
  ggsave("../figures/SKMEL5_allClusters_CellCycleState_proportion_updated20220129.pdf", width = 6, height = 5)
```

#### Figure 2B original
```{r}
####

EMD_data_UT <- subset(FetchData(combined, vars = c("UMAP_1", "UMAP_2", "old.ident", "BCnum", "lineageColored")), 
                      old.ident == "Untreated")
D_UT <- dist(cbind(EMD_data_UT$UMAP_1, EMD_data_UT$UMAP_2), 
          diag=TRUE, upper=TRUE) 
m_UT <- as.matrix(D_UT) # coerce dist object to a matrix
dimnames(m_UT) <- list(1:length(EMD_data_UT$UMAP_1),
                    1:length(EMD_data_UT$UMAP_2))
xy_UT <- t(combn(colnames(m_UT), 2))
df_UT_EMD <- data.frame(xy_UT, dist=m_UT[xy_UT])
df_UT_EMD$Condition <- "Untreated"

EMD_data_I <- subset(FetchData(combined, vars = c("UMAP_1", "UMAP_2", "old.ident", "BCnum", "lineageColored")), 
                      old.ident == "Idling")
D_I <- dist(cbind(EMD_data_I$UMAP_1, EMD_data_I$UMAP_2), 
             diag=TRUE, upper=TRUE) 
m_I <- as.matrix(D_I) # coerce dist object to a matrix
dimnames(m_I) <- list(1:length(EMD_data_I$UMAP_1),
                       1:length(EMD_data_I$UMAP_2))
xy_I <- t(combn(colnames(m_I), 2))
df_I_EMD <- data.frame(xy_I, dist=m_I[xy_I])
df_I_EMD$Condition <- "Idling"

df_all_EMD <- rbind(df_UT_EMD, df_I_EMD)

df_all_EMD_UT <- df_UT_EMD[sample(nrow(df_UT_EMD), 15000), ]
df_all_EMD_I <- df_I_EMD[sample(nrow(df_I_EMD), 15000), ]
df_all_EMD_sample <- rbind(df_all_EMD_UT, df_all_EMD_I)

EMD_compare <- wasserstein1d(df_all_EMD_UT$dist, df_all_EMD_I$dist)

df_all_EMD_sample$Condition <- factor(df_all_EMD_sample$Condition,
                                      levels = c("Untreated", "Idling"))
ggplot(df_all_EMD_sample, aes(dist, color = Condition)) + 
  theme_bw() + stat_ecdf(size = 1.25) +
  scale_color_manual(values = c("red", "blue")) +
  labs(x = "Distance", y = "Fraction") +
  annotate("text", x = 17.5, y = 0.2, label = paste("EMD =", round(EMD_compare,2)), size = 5) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        axis.text.x = element_text(size = 12, colour = "black"),
        axis.text.y = element_text(size = 12, colour = "black"),
        legend.position = "none", legend.text = element_text(size = 12),
        plot.title = element_text(size = 12, hjust = 0.5, face = "bold"), axis.text=element_text(size=14),
        legend.title = element_text(size=12), axis.title=element_text(size=12)) 
  ggsave("../figures/ECDF_UT-I_withEMD.pdf", width = 4, height = 3)

```

Instead of using UMAP and euclidean distance, I'm going to try doing a similar thing using the first 10 PCs. 

```{r}
# same process as above, except extracting the PC data
EMD_data_UT <- subset(FetchData(combined, vars = c("PC_1", "PC_2", "PC_3", "PC_4", "PC_5", "PC_6", "PC_7", "PC_8", "PC_9", "PC_10", "old.ident", "BCnum", "lineageColored")),
                      old.ident == "Untreated")
D_UT <- dist(cbind(EMD_data_UT$PC_1, EMD_data_UT$PC_2, EMD_data_UT$PC_3, EMD_data_UT$PC_4, EMD_data_UT$PC_5, EMD_data_UT$PC_6, EMD_data_UT$PC_7, EMD_data_UT$PC_8, EMD_data_UT$PC_9, EMD_data_UT$PC_10), diag=TRUE, upper=TRUE)
m_UT <- as.matrix(D_UT) # coerce dist object to a matrix
dimnames(m_UT) <- list(1:length(EMD_data_UT$PC_1),
                    1:length(EMD_data_UT$PC_2))
xy_UT <- t(combn(colnames(m_UT), 2))
df_UT_EMD <- data.frame(xy_UT, dist=m_UT[xy_UT])
df_UT_EMD$Condition <- "Untreated"

EMD_data_I <- subset(FetchData(combined, vars = c("PC_1", "PC_2", "PC_3", "PC_4", "PC_5", "PC_6", "PC_7", "PC_8", "PC_9", "PC_10", "old.ident", "BCnum", "lineageColored")),
                      old.ident == "Idling")
D_I <- dist(cbind(EMD_data_I$PC_1, EMD_data_I$PC_2, EMD_data_I$PC_3, EMD_data_I$PC_4, EMD_data_I$PC_5, EMD_data_I$PC_6, EMD_data_I$PC_7, EMD_data_I$PC_8, EMD_data_I$PC_9, EMD_data_I$PC_10), 
             diag=TRUE, upper=TRUE) 
m_I <- as.matrix(D_I) # coerce dist object to a matrix
dimnames(m_I) <- list(1:length(EMD_data_I$PC_1),
                       1:length(EMD_data_I$PC_2))
xy_I <- t(combn(colnames(m_I), 2))
df_I_EMD <- data.frame(xy_I, dist=m_I[xy_I])
df_I_EMD$Condition <- "Idling"

df_all_EMD <- rbind(df_UT_EMD, df_I_EMD)

df_all_EMD_UT <- df_UT_EMD[sample(nrow(df_UT_EMD), 15000), ]
df_all_EMD_I <- df_I_EMD[sample(nrow(df_I_EMD), 15000), ]
df_all_EMD_sample <- rbind(df_all_EMD_UT, df_all_EMD_I)

EMD_compare <- wasserstein1d(df_all_EMD_UT$dist, df_all_EMD_I$dist)

df_all_EMD_sample$Condition <- factor(df_all_EMD_sample$Condition,
                                      levels = c("Untreated", "Idling"))

ggplot(df_all_EMD_sample, aes(dist, color = Condition)) + 
  theme_bw() + stat_ecdf(size = 1.25) +
  scale_color_manual(values = c("red", "blue")) +
  labs(x = "Distance", y = "Fraction") +
  annotate("text", x = 37.5, y = 0.2, label = paste("EMD =", round(EMD_compare,2)), size = 5) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        axis.text.x = element_text(size = 12, colour = "black"),
        axis.text.y = element_text(size = 12, colour = "black"),
        legend.position = "none", legend.text = element_text(size = 12),
        plot.title = element_text(size = 12, hjust = 0.5, face = "bold"), axis.text=element_text(size=14),
        legend.title = element_text(size=12), axis.title=element_text(size=12)) 
  # ggsave("../figures/ECDF_UT-I_withEMD_PCA-10PC.pdf", width = 4, height = 3)
```


I want to find the total variance explained by the PCA analysis above, but this is very difficult given the state of the object. Therefore, I did the following to get the variance of the object and the eigenvalues to calculate the variance explained by each PC. I got this solution from the following message board on github: https://github.com/satijalab/seurat/issues/982
```{r}
mat <- Seurat::GetAssayData(combined, assay = "SCT", slot = "scale.data")
pca <- combined[["pca"]]
total_variance <- sum(matrixStats::rowVars(mat))

eigValues <-(pca@stdev)^2  ## EigenValues
varExplained <- eigValues / total_variance

sum(varExplained)
```
For the combined, combined_untreated, and combined_idling seurat objects, their first 50 PCs only explain 0.0363, 0.0358, and 0.0381 of the variance in the data, respectively. This seems really low, so we may want to consult with experts on this. That doesn't mean that our results are not meaningful, but there may be some things we can do to have a stronger analysis. 

